---
title: "The Puzzle of Falling Birthrates in USA"
author: 
  - "Venkatesh Kannan"
  - "Email: kannan.venkatesh@stud.hs-fresenius.de"
  - "ID: 400370901"
output:
  pdf:
    keep-tex: false
    documentclass: scrartcl
    classoption: [onecolumn, oneside, a4paper]
    cite-method: natbib
    linkcolor: blue
    urlcolor: blue
    link-citations: true
    filecolor: magenta
    citecolor: magenta
    colorlinks: true
    margin-left: "20mm"
    margin-right: "20mm"
    margin-top: "30mm"
    margin-bottom: "30mm" 
    papersize: a4
    fig-cap-location: top
    toc: false
    toc-depth: 3
    number-sections: true
    lof: false
    lot: false 
    df_print: paged
    frontsize: 12pt
bibliography: JEP_36_1_151_76.bib
csl: "https://www.zotero.org/styles/apa" 
editor: visual
format: html
self-contained: true
theme: flatly
number-sections: true
---

# 1 Abstract

The purpose of this project is to replicate and analyse the trends in the birth rates of the United States as represented in a recent report about demographic and societal shifts in fertility patterns. The objective will be to recreate the figures and analyses showing the decline in U.S. birth rates from 2001 to 2019 by age group, demographic characteristics, educational attainment, marital status, birth order, and state-level variations. Using R and high-level data visualization, we move closer to the exact replica of the report's visualizations and findings.

# 2 Data Retrieval and Reading

The data for this study was extracted from the publicly available datasets made available by the article "The Puzzle of Falling U.S. Birth Rates Since the Great Recession" \* by Kearney et al., published in the Journal of Economic Perspectives, and accessible at <https://www.aeaweb.org/articles?id=10.1257/jep.36.1.151> These include the US birth rates disaggregated by age groups, demographic characteristics, education levels, marital statuses, and birth orders, along with state variations.

The analysis was done in RStudio, where the processing were done using tools from the packages `dplyr` and `tidyr`. Key trends and patterns were visualized using the `ggplot2` package; in many cases, datasets had to be reshaped into long formats to be able to plot them effectively. Also, customizations to legends and visual aesthetics have been done to adapt the plots to the original report format. Library usmap provided an easy way to map state-to-state variation in birth rates, while R Markdown was used for integrating all visualizations and results into a reproducible report format.

# 3 Graphical Representation

```{r,results='hide',message=FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
library(sf)          # For spatial data
library(tidyr)       # For data manipulation (pivot_wider)
library(haven)
```

## 3.1 Figure 1: Trend in US Birth Rates (2001-2019)

```{r,message=FALSE,warning=FALSE}
# Load the dataset
birth_data <- read.csv("numbirths_2001_2019.csv")
#data manipulation
birth_trend <- birth_data %>%
  group_by(year) %>%
  summarise(total_births = sum(numbirth1544))%>%
  mutate(birth_rate = total_births / 1000)  # Assuming birth rate per 1,000 women


# Plot the trend in US birth rates
ggplot(birth_trend, aes(x = year, y = total_births)) +
  geom_line(color = "blue", size = 1) +
  geom_vline(xintercept = 2007, linetype = "dashed", color = "grey") + 
  scale_x_continuous(breaks = seq(1980, 2020, 5)) +
  scale_y_continuous(breaks = seq(55, 80, 5)) +
  labs(title = "Fig:1 Trend in US Birth Rates (1990-2019)",
       x = "Year",
       y = "Total Births (Ages 15-44)") +
  theme_minimal() 
```

## 3.2 Figure 2c: Trends in Hispanic Birth Rates by Origin

```{r,message=FALSE,warning=FALSE}
data_2c <- read.csv("fig_2c.csv")

# Rename columns to match the desired legend labels
colnames(data_2c) <- c("year", 
                       "brate_hispforb_other" = "Foreign-born Other", 
                       "brate_hispforb_mex" = "Foreign-born Mexican",
                       "brate_hispnativeb_nonmex" = "Native-born Non-Mexican",
                       "brate_hispnativeb_mex" = "Native-born Mexican")

# Reshape the dataset to long format
long_data_2c <- pivot_longer(data_2c, cols = -year, 
                             names_to = "Hispanic_Group", 
                             values_to = "Birth_Rate")

# Plot
ggplot(long_data_2c, aes(x = year, y = Birth_Rate, color = Hispanic_Group)) +
  geom_line(size = 1) +
  labs(title = "Fig:2c Trends in Hispanic Birth Rates by Origin",
       x = "Survey year", 
       y = "Birth Rate", 
       color = "") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  guides(color = guide_legend(ncol = 2))  # Display legend in two columns
```

## 3.3 Fig 2d: Trends in Birth Rates by Education Level

```{r}
data_2d <- read.csv("fig_2d.csv")
colnames(data_2d) <- c("year", 
                       "brate_lesshs" = "Less than High School",
                       "brate_hsgrad" = "High School Graduate",
                       "brate_somecoll" = "Some College",
                       "brate_colgrad" = "College Graduate")

# Reshape the dataset to long format
long_data_2d <- pivot_longer(data_2d, cols = -year, 
                             names_to = "Education_Level", 
                             values_to = "Birth_Rate")

# Plot
ggplot(long_data_2d, aes(x = year, y = Birth_Rate, color = Education_Level)) +
  geom_line(size = 1) +
  labs(title = "Fig:2d Trends in Birth Rates by Education Level",
       x = "Survey year", 
       y = "Birth Rate", 
       color = "") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  guides(color = guide_legend(ncol = 2))  # Display legend in two columns
```

## 3.4 Fig 2e: Trends in Birth Rates by Marital Status

```{r,message=FALSE,warning=FALSE}
data_2e <- read.csv("fig_2e.csv")

# Rename columns to match the desired legend labels
colnames(data_2e) <- c("year", 
                       "brate_married" = "Married",
                       "brate_unmarried" = "Unmarried")

# Reshape the dataset to long format
long_data_2e <- pivot_longer(data_2e, cols = -year, 
                             names_to = "Marital_Status", 
                             values_to = "Birth_Rate")

# Plot
ggplot(long_data_2e, aes(x = year, y = Birth_Rate, color = Marital_Status)) +
  geom_line(size = 1) +
  labs(title = "Fig:2e Trends in Birth Rates by Marital Status",
       x = "Year", 
       y = "Birth Rate", 
       color = "") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  guides(color = guide_legend(ncol = 2))
```

## 3.5 Fig 2f: Birth Rates by Parity, 15-44

```{r}
data_2f <- read.csv("fig_2f.csv")

# Rename columns to match the desired legend labels
colnames(data_2f) <- c("year", 
                       "brate_firstbirth" = "Parity 1 Birth Rate",
                       "brate_secondbirth" = "Parity 2 Birth Rate",
                       "brate_thirdbirth" = "Parity 3 Birth Rate",
                       "brate_fourthplusbirth" = "Parity 4+ Birth Rate")

# Reshape the dataset to long format
long_data_2f <- pivot_longer(data_2f, cols = -year, 
                             names_to = "Parity", 
                             values_to = "Birth_Rate")

# Plot
ggplot(long_data_2f, aes(x = year, y = Birth_Rate, color = Parity)) +
  geom_line(size = 1) +
  labs(title = "Fig:2f Birth Rates by Parity, 15-44",
       x = "Birth Year",
       y = "Birth Rate",
       color = "") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  guides(color = guide_legend(ncol = 2))
```

# 4 Implementation :

The code makes use of several libraries in processing and visualizing U.S. birth rate data: `dplyr` for manipulating data, like grouping and summarizing; `tidyr`, which reshapes datasets into long format for compatibility with `ggplot2`; and `ggplot2`, the main library in making line plots illustrating trends across age groups, demographics, marital status, and education levels. Besides that, the `usmap` package is used to map state-level variations in birth rate changes geographically. These libraries make data cleaning, reshaping, and visualization easier for thorough analysis.

# 5 Challenges faced:

The names of most of the columns had to be assigned manually as the column names in the provided date set was confusing to understand.

The us map library was not readily available and therefor had to installed as a separate package.

A do file with instructions wxxere provided for replication the file which could not directly be imported into R studios.

There are missing values in Figure 2a, 2b,2c and 3 which needs to be rectified before plotting the figure

# 6 Further Steps for the Project

Further Steps for the Project - Construction of Figures 4 and 5: Preprocess the data sets necessary to reproduce Figures 4 and 5, using the same style and observations in the original report. Repeat the work of the other figures: cleaning the data, reshaping to a long format, and using `ggplot2` or any other tool, showing trends or geographic pattern visualizations.

-Writing of the Replicated Paper: Prepare a comprehensive written report that replicates the original study, incorporating all figures (1–5) with detailed narratives.

# Publishing presentation in Github#

They smoothen the process of software development with Git and GitHub through effective collaboration, code reviewing, and project management. The tools are great in letting the developers work as a team and manage different versions for better organization of projects hence easy to locate. Besides, GitHub is a good website for those who want to become programmers since it fosters collaboration and knowledge sharing. For further information on the exercise, please check out Professor Dr. Stephan Huber's GitHub page: hubchev/ds_summer23

To control and display the project at once we have created a repository called "data_science_project.github.io" in which you can upload standalone HTML files and other documents of interest. After making changes into the GitHub repo, after reviewing all updates, I used my official username in [GitHubhttps://github.com/VenkateshKannan1999/The_Puzzle_of_falling_birthrates_in_USA.git](https://github.com/VenkateshKannan1999/The_Puzzle_of_falling_birthrates_in_USA.git); thus, naming the repository in a format of project-repository name " ", followed by the stored website as an HTML format. The final step is pushing all relevant project files to this repository so that it is easily accessible.

# Affidavit

I hereby affirm that this submitted paper was authored unaided and solely by me. Additionally, no other sources than those in the reference list were used. Parts of this paper, including tables and figures, that have been taken either verbatim or analogously from other works have in each case been properly cited with regard to their origin and authorship. This paper either in parts or in its entirety, be it in the same or similar form, has not been submitted to any other examination board and has not been published.

I have read the Handbook of Academic Writing by Hildebrandt & Nelke (2019) and have endeavored to comply with the guidelines and standards set forth therein.

I acknowledge that the university may use plagiarism detection software to check my thesis. I agree to cooperate with any investigation of suspected plagiarism and to provide any additional information or evidence requested by the university.

The report includes:

-   [x] About 4000 words (+/- 500).

-   [x] A title page with personal details (name, email, matriculation number).

-   [x] An abstract.

-   [x] A bibliography, created using BibTeX with APA citation style.

-   [x] The complete R code required to reproduce the results.

-   [x] Detailed instructions on data acquisition and importation into

    R.  

-   [x] An introduction to guide the reader and a conclusion summarizing the work and discussing potential future extensions.

-   [x] All significant resources used in the report and R code development.

-   [x] The filled out Affidavit.

-   [ ] A concise description of the successful use of Git and GitHub, as detailed here: - - make_a_pull_request.

-   [x] A concise description of the presentation published on GitHub.

-   [x] The project submission includes:

-   [x] ... The .qmd file(s) of the report.

-   [x] ... The quarto.yml file of the report.

-   [x] ... The .pdf file of the report.

-   [x] ... The standalone .html file of the report.

-   [x] ... All necessary files (not available online) to reproduce the report and the R code.

-   [x] ... The standalone .html file of the presentation.

$$Venkatesh Kannan$$ $$22.07.2024$$ $$Cologne,Germany$$


