---
title: 'Reproducing the visualisations using R programming language'
author:
- "Email: ravanasamuthram_krishnan.pooja@stud.hs-fresenius.de"
- 'Name: Pooja Ravanasamuthram Krishnan'
- 'ID: 400376705'
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    fig_caption: true
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: '3'
    number_sections: true
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
header-includes: \usepackage{setspace}\doublespacing
fontsize: 12pt
urlcolor: blue
linkcolor: red
bibliography: "references.bib"
csl: "https://www.zotero.org/styles/apa"
editor_options:
  markdown:
   wrap: 72
---

# **Abstract**

This paper analyzes the origin and trends of economic sanctions (in
international law also referred to as restrictive measures), 1950–2022,
drawing from the Global Sanctions Database. We reproduce the graphs in
the study using the R programming language (original code provided in
Stata). The dataset needed for the graphs is however not provided, which
creates a problem for replication. The consequences of sanctions on
political goals are central to this paper. The types of graphs are line
graph, bar graph, scatter plot, stacked area chart and composite graph
showing different trend in data. As observed, despite the overall
complexity and variety of visual trends present in the data, all graphs
have similarities in the high customisation in order to further make the
data easily understandable, the use of the same colour schemes to
distinguish different categories, and the concentration of time as a
common factor as the X-axis, illustrating trends and shifts over
decades. This project demonstrates the utility of R in replicating
complex visualizations and offers insights into the process of analyzing
economic sanctions data.

# **Project Overview**

The project explores converting raw data into meaningful visualizations
through R, a statistical programming language, in terms of how well they
reflect findings in the paper. This project is also quite useful in
understanding how the methodologies were done in the study while it has
practical lessons on the challenges and fine points of data analysis and
visualization in R.This project focuses on replicating visualizations
from the paper "Economic Sanctions: Evolution, Consequences, and
Challenges." The paper discusses the history, objectives, and
effectiveness of economic sanctions, with an emphasis on their evolution
over time and their political and economic goals.

# **Significance of using R**

R is widely used in the data science job market and can be run on any
operating system, such as Windows, macOS, Linux, and many other
environments. It can be integrated with various programming languages,
including C, C#, and Python. The number of R users and those interested
in learning it is significantly increasing, largely because it is free
to use, eliminating the need for financial investment and licensing. As
an open-source programming language, R is accessible to anyone. With the
growing number of users, many skilled professionals use R as part of
their profession, while others learn it out of personal interest.

# **Initial steps for replication**

The capabilities of this language are extended by the Comprehensive R
Archive Network (CRAN), which contains numerous libraries and packages,
such as ggplot2 and dplyr. These resources can be applied in various
fields, including machine learning and bioinformatics, further enhancing
the versatility and utility of R. Necessary packages are installed to
manipulate the data for analysis and visualization.

```{r}

options(repos = c(CRAN = "https://cran.rstudio.com/"))

install.packages("dplyr")

install.packages("haven")

install.packages("ggplot2")

install.packages("reshape2")


```

To reproduce the graphs, which are stacked area graph and bar charts,
necessary libraries such as 'dplyr', 'tidyr', 'ggplot2', and 'readxl'
are loaded. Since the dataset is in Excel format, the 'readxl' library
is used as it supports older Excel file formats like '.xls' and '.xlsx'.
The Excel file is loaded by mentioning the path of the file, and the
sheet = 1 argument is specified to indicate that the first sheet of the
Excel file should be loaded. And I have shown how the original dataset
looks like. Later, for the purpose of reproducing the graphs, I have
manipulated the dataset.

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)
library(tidyverse)


gsdb_data <- read_excel("C:/Users/pooja/OneDrive/Data Science- Economic Sanctions/GSDB_V3/GSDB_V3.xls", sheet = 1)

head(gsdb_data)

```

### **Replicating graph 1:**

In the dataset, the column 'other' is renamed to 'other_sanct' for
clarity and to avoid potential conflicts, as 'other' is too generic. The
'objective' column is divided into four separate columns: 'objective1',
'objective2', 'objective3', and 'objective4'.This is because the dataset
uses commas to store up to four values in a single row. A for loop is
used which gives an output value of 1 if a specific objective exists in
any of the four split columns and 0 if it does not. Similarly, the
'success' column is split into four columns, and the same procedure is
applied to the 'success' column.

```{r}

gsdb_data <- gsdb_data %>% rename(other_sanct = other)

gsdb_data <- gsdb_data %>%
  separate(objective, into = c("objective1", "objective2", "objective3", 
  "objective4"), sep = ",", fill = "right")

objectives <- c("democracy", "destab_regime", "end_war", "human_rights", 
                "terrorism", "territorial_conflict", "prevent_war", 
                "policy_change", "other")

for (obj in objectives) {
  gsdb_data[[obj]] <- ifelse(gsdb_data$objective1 == obj | gsdb_data$objective2
                             == obj | gsdb_data$objective3 == obj | 
                               gsdb_data$objective4 == obj, 1, 0)
}

gsdb_data <- gsdb_data %>%
  separate(success, into = c("success1", "success2", "success3", "success4"),
           sep = ",", fill = "right")

success_types <- c("success_total", "success_part", "nego_settlement", "failed",
                   "ongoing")

for (succ in success_types) {
  gsdb_data[[succ]] <- ifelse(gsdb_data$success1 == succ | gsdb_data$success2 ==
                                succ | gsdb_data$success3 == succ | 
                                gsdb_data$success4 == succ, 1, 0)
}

```

Two datasets are created: sanct_begin and sanct_end and two columns are
selected for each of the dataset. For the sanct_begin dataset, the
columns case_id and begin are selected, with the begin column renamed to
year. Similarly, for the sanct_end dataset, the columns case_id and end
are selected, with the end column renamed to year. Both datasets are
then integrated (bound together), duplicates are removed, and only cases
from the year 1950 and onward are retained. For each case_id, the begin
and end years are marked, and all the years between the begin and end
years are filled. For example, if the begin year for a case_id is 1950
and the end year is 1954, it is expanded to include 1950, 1951, 1952,
1953, and 1954. Additional columns, such as objective and success, are
incorporated into the expanded dataset using the left_join function. A
mutate function is applied to the expanded dataset to add a new column,
id_new. This column outputs 1 if the year corresponds to the starting
year of a sanction case and 0 otherwise. Finally, for each year, the
total number of cases (total_cases) and new cases (new_cases) are
calculated, excluding rows with missing values in the id_new column.

```{r}

sanct_begin <- gsdb_data %>%
  select(case_id, begin) %>%
  rename(year = begin)

sanct_end <- gsdb_data %>%
  select(case_id, end) %>%
  rename(year = end)

gsdb_case <- bind_rows(sanct_begin, sanct_end) %>%
  distinct() %>%
  arrange(case_id, year) %>%
  filter(year >= 1950) 

expanded_data <- gsdb_case %>%
  group_by(case_id) %>%
  complete(year = full_seq(year, 1)) %>%
  left_join(gsdb_data, by = "case_id")

expanded_data <- expanded_data %>%
  filter(year >= 1950)

expanded_data <- expanded_data %>%
  mutate(id_new = ifelse(year %in% sanct_begin$year[sanct_begin$case_id == 
  case_id], 1, 0))

expanded_data <- expanded_data %>%
  mutate(id = 1)

collapsed_data <- expanded_data %>%
  group_by(year) %>%
  summarise(total_cases = sum(id, na.rm = TRUE),
            new_cases = sum(id_new, na.rm = TRUE))

```

Finally, using ggplot and ggsave function, the graph is plotted and
saved.

```{r}


ggplot(collapsed_data, aes(x = year)) +
  geom_area(aes(y = total_cases, fill = "Total Cases"), alpha = 0.7) +  
  geom_area(aes(y = new_cases, fill = "New Cases"), alpha = 0.7) +      
  scale_fill_manual(values = c("Total Cases" = "blue", "New Cases" = "red")) +
  labs(y = "Number of sanction cases", x = "Years") +
  scale_y_continuous(breaks = seq(0, 400, by = 50)) +
  scale_x_continuous(breaks = seq(1950, 2022, by = 10)) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top", legend.title = element_blank())

ggsave("evol_sanct.png", width = 8, height = 6, dpi = 300)
ggsave("figure_1.pdf", width = 8, height = 6)

```

### **Replicating graph 2:**

To replicate the 2nd and 3rd graphs, I used the same procedure of
renaming the begin and end columns, integrating them and dedepulicating
the redundant data, and considering the case_id for all the years
starting from 1950. The new dataset with these changes is assigned to a
different dataset named sanction_summary, and it is grouped by year.
Then, the summarize function is used to sum the values of all the
columns (arms, trade, travel, military, financial, other_sanct) by
removing the 'NA' values. The mutate function is then used to create the
columns sum2, sum3, sum4, sum5, and sum6.

```{r}

sanct_begin <- gsdb_data %>%
  select(case_id, begin) %>%
  rename(year = begin)

sanct_end <- gsdb_data %>%
  select(case_id, end) %>%
  rename(year = end)

gsdb_case <- bind_rows(sanct_begin, sanct_end) %>%
  distinct() %>%
  arrange(case_id, year) %>%
  filter(year >= 1950)

expanded_data <- gsdb_case %>%
  group_by(case_id) %>%
  complete(year = full_seq(year, 1)) %>%
  left_join(gsdb_data, by = "case_id") %>%
  filter(year >= 1950)

expanded_data <- expanded_data %>%
  mutate(
    id_new = ifelse(year %in% sanct_begin$year[sanct_begin$case_id == case_id], 1, 0),
    id = 1
  )

sanction_summary <- expanded_data %>%
  group_by(year) %>%
  summarise(
    arms = sum(arms, na.rm = TRUE),
    military = sum(military, na.rm = TRUE),
    trade = sum(trade, na.rm = TRUE),
    financial = sum(financial, na.rm = TRUE),
    travel = sum(travel, na.rm = TRUE),
    other_sanct = sum(other_sanct, na.rm = TRUE)
  ) %>%
  mutate(
    sum2 = arms + military,
    sum3 = arms + military + trade,
    sum4 = arms + military + trade + financial,
    sum5 = arms + military + trade + financial + travel,
    sum6 = arms + military + trade + financial + travel + other_sanct
  )

```

The mutate function was used again to create six new columns to properly
align the graph for different sanction types. Then, only the columns
required for plotting the graph were included using the select function.
To group the sanction types under one column and their counts under
another, the pivot_longer function was used. For ordering purposes in
the graph, the levels and labels were adjusted using mutate. Finally,
colors were assigned to each sanction type, and the graph was plotted.

```{r}

plot_data <- sanction_summary %>%
  mutate(
    inc_arms = arms,
    inc_military = sum2 - arms,
    inc_trade = sum3 - sum2,
    inc_financial = sum4 - sum3,
    inc_travel = sum5 - sum4,
    inc_other = sum6 - sum5
  ) %>%
  select(year, inc_arms, inc_military, inc_trade, inc_financial, inc_travel, inc_other) %>%
  pivot_longer(
    cols = starts_with("inc_"),
    names_to = "category",
    values_to = "value"
  ) %>%
  mutate(
    category = factor(
      category,
      levels = rev(c("inc_arms", "inc_military", "inc_trade", "inc_financial", "inc_travel", "inc_other")),
      labels = rev(c("Arms", "Military", "Trade", "Financial", "Travel", "Other"))
    )
  )

colors <- c(
  "Arms" = "blue", 
  "Military" = "maroon", 
  "Trade" = "forestgreen", 
  "Financial" = "darkorange", 
  "Travel" = "grey", 
  "Other" = "red"
)

ggplot(plot_data, aes(x = year, y = value, fill = category)) +
  geom_area(alpha = 0.8, position = "stack", color = "black") +  
  scale_fill_manual(values = colors) +
  labs(
    y = "Number of sanctions by type",
    x = "Years",
    title = "Number of Sanctions by Type Over Time"
  ) +
  scale_y_continuous(breaks = seq(0, 900, by = 100), limits = c(0, 900), expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(1950, 2022, by = 10), expand = c(0, 0)) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.background = element_rect(fill = "white", color = "white")
  )

```

### **Replicating graph 3:**

For plotting the graph, six columns are created, each column containing
the cumulative percentage of sanction types added consecutively. Perc6
is assigned the value 1 because it represents the total of all sanction
types divided by itself, resulting in a value of 1. Next, the inc1,
inc2, inc3, inc4, inc5, and inc6 columns are created, with each column
showing the difference between consecutive perc values calculated in the
previous step. Using these values, the graph is plotted and saved.

```{r}

sanction_summary <- sanction_summary %>%
  mutate(
    perc1 = arms / (arms + military + trade + financial + travel + other_sanct),
    perc2 = (arms + military) / (arms + military + trade + financial + travel + other_sanct),
    perc3 = (arms + military + trade) / (arms + military + trade + financial + travel + other_sanct),
    perc4 = (arms + military + trade + financial) / (arms + military + trade + financial + travel + other_sanct),
    perc5 = (arms + military + trade + financial + travel) / (arms + military + trade + financial + travel + other_sanct),
    perc6 = 1,  
    inc1 = perc1,
    inc2 = perc2 - perc1,
    inc3 = perc3 - perc2,
    inc4 = perc4 - perc3,
    inc5 = perc5 - perc4,
    inc6 = perc6 - perc5
  )

plot_data <- sanction_summary %>%
  select(year, inc1, inc2, inc3, inc4, inc5, inc6) %>%
  pivot_longer(cols = starts_with("inc"), 
               names_to = "category", values_to = "value") %>%
  mutate(
    
    category = factor(category, 
                      levels = rev(c("inc1", "inc2", "inc3", "inc4", "inc5", "inc6")),
                      labels = rev(c("Arms", "Military", "Trade", "Financial", "Travel", "Other")))
  )

colors <- c("Arms" = "navy", 
            "Military" = "maroon", 
            "Trade" = "forestgreen", 
            "Financial" = "darkorange", 
            "Travel" = "grey", 
            "Other" = "red")


ggplot(plot_data, aes(x = year, y = value, fill = category)) +
  geom_area(alpha = 0.7) +  
  scale_fill_manual(values = colors) +  
  labs(
    y = "Percent of sanctions by type",
    x = "Years",
    title = "Percent of Sanctions by Type Over Time"
  ) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.2), limits = c(0, 1), expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(1950, 2022, by = 10), expand = c(0, 0)) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.background = element_rect(fill = "white", color = "white")
  )
ggsave("evol_type_perc.png", width = 10, height = 6, dpi = 300)
ggsave("figure_2b.pdf", width = 10, height = 6)

```

### **Replicating graph 4:**

For plotting the 4th graph, a new dataset is created with specific
columns such as Objective, Category1, and Category2. Category1 and
Category2 contain percentage values for the objectives. By default,
categorical variables are unordered or ordered alphabetically. If a
specific order is required, the factor function can be used. Finally, to
plot the graph, the categories are aggregated into a single column using
the pivot_longer function.

```{r}

data <- data.frame(
  Objective = c("Terrorism", "Territorial conflict", "Destabilize regime", 
                "Policy change", "Other", "Prevent war", "End war", 
                "Human rights", "Democracy"),
  Category1 = c(72, 62, 52, 48, 43, 42, 38, 30, 15),
  Category2 = c(28, 38, 48, 52, 57, 58, 62, 70, 85)  
)

data$Objective <- factor(data$Objective, 
                         levels = c("Terrorism", "Territorial conflict", 
                                    "Destabilize regime", "Policy change", 
                                    "Other", "Prevent war", "End war", 
                                    "Human rights", "Democracy"))

data_long <- data %>%
  tidyr::pivot_longer(cols = c("Category1", "Category2"), 
                      names_to = "Category", values_to = "Percent")

ggplot(data_long, aes(x = Percent, y = Objective, fill = Category)) +
  geom_bar(stat = "identity", position = "stack", color = "white") +
  scale_fill_manual(values = c("Category1" = "brown", "Category2" = "steelblue")) +
  scale_x_continuous(breaks = seq(0, 100, by = 20)) +  
  labs(x = "Percent", y = NULL, title = "Panel A. By objective") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )


```

# **Learning with R language**

R is a completely new programming language for me. When I first learned
about the course that included R, it sparked my interest. By attending
Prof. Huber's lectures and following his notes, I was able to gain a
solid understanding of the functions and packages used in R, as well as
the way it be manipulated to achieve the desired results. In addition, I
used online platforms to learn and deepen my comprehension of various
functions. Although I still have a long way to go in this learning
process, I can now confidently say that I know how to use the tools
required to code in the R programming language.

I recreated four graphs in my report, as the first three are quite
similar. I have now learned how to reproduce similar graphs with varied
data. The last one is a simple bar graph. Throughout the process of
replicating these graphs, I learned more functions and their purposes.
In this report, I have made my best effort to explain the code I wrote
with the help of ChatGPT and online platforms. The provided code was
given in Stata and it was difficut to understand.

# **Learning about Git and Github**

Git and GitHub are widely used in software development projects. They
help track changes in the code, facilitate sharing, and enable
collaboration with others. This is particularly useful when you have
written a piece of code and need someone to review it. Even though I had
basic knowledge of this software, I learned more about it for this
particular project with the help of my professor's tutorials and the
step-by-step procedure that he shared in his Github account. It turned
out to be very beneficial, as this software is widely used in
professional environments.

I used GitHub to share my progress and my final report, and I also
created a pull request with the help of the tutorial my professor shared
with me. Specifically, I created a repository to upload my work and
forked my professor's repository. Then, I cloned the forked repository
to my local machine, made changes to it (to share my work for his
review), and pushed the changes back to my forked repository. Finally, I
created a pull request from my forked repository to my professor's
repository.

# **Challenges Encountered**

I faced difficulties while replicating the graphs, especially the
stacked area graphs, as the shaded areas initially got overlapped, and I
didn't know how to arrange them accordingly. To resolve this problem, I
subtracted the area which overlapped from the total and stack each layer
sequentially for proper alignment.

# **Conclusion and future work**

It was a great experience for me working on this paper. By attending
classes and clearing exams, I would have learned the syntax and
functions of the R language, but by reproducing a graph that was coded
in a different programming language, I learned how to interpret the
logic between programming languages and troubleshoot the issues.

In the future, I will make efforts to try replicating different types of
graphs, learning complex functionalities and algorithms. I will strive
to code from scratch, explore advanced techniques, and stay updated with
the latest developments in the R environment to enhance my skills
further.

# **Affidavit**

I hereby affirm that this submitted paper was authored unaided and
solely by me. Additionally, no other sources than those in the reference
list were used. Parts of this paper, including tables and figures, that
have been taken either verbatim or analogously from other works have in
each case been properly cited with regard to their origin and
authorship. This paper either in parts or in its entirety, be it in the
same or similar form, has not been submitted to any other examination
board and has not been published.

I have read the Handbook of Academic Writing by Hildebrandt & Nelke
(2019) and have endeavored to comply with the guidelines and standards
set forth therein.

I acknowledge that the university may use plagiarism detection software
to check my thesis. I agree to cooperate with any investigation of
suspected plagiarism and to provide any additional information or
evidence requested by the university.

The report includes:

-   [x] About 4000 words (+/- 500).

-   [x] A title page with personal details (name, email, matriculation
    number).

-   [x] An abstract.

-   [x] A bibliography, created using BibTeX with APA citation style.

-   [x] The complete R code required to reproduce the results.

-   [x] Detailed instructions on data acquisition and importation into

    R.  

-   [x] An introduction to guide the reader and a conclusion summarizing
    the work and discussing potential future extensions.

-   [x] All significant resources used in the report and R code
    development.

-   [x] The filled out Affidavit.

-   [x] A concise description of the successful use of Git and GitHub,
    as detailed here: - - make_a_pull_request.

-   [x] A concise description of the presentation published on GitHub.

-   [x] The project submission includes:

-   [x] ... The .qmd file(s) of the report.

-   [x] ... The \_quarto.yml file of the report.

-   [x] ... The .pdf file of the report.

-   [x] ... The standalone .html file of the report.

-   [x] ... All necessary files (not available online) to reproduce the
    report and the R code.

-   [x] ... The standalone .html file of the presentation.

Pooja Ravanasamuthram Krishnan, 03.02.2025, Cologne, Germany

```{r, echo=TRUE, results='asis'}

text <- readLines("Final_report_DS2025.qmd")  
text <- text[!grepl("^---", text)]
word_count <- sum(sapply(strsplit(text, "\\s+"), length))
cat(paste("Word count:", word_count))

```


# **References**
1. @10.1257/jep.37.1.3